signalHsmm - a novel semi-Markov model of eukaryotic signal peptides

### 2015-06-17 paper.tex
adding info in signalHSMM diagram

\subsection*{Hidden semi-Markov model}
Hidden semi-Markov model (HSMM) is an extension of popular method  hidden Markov model (HMM).
In many fields it proved to yield better accuracy of bioinformatics algorithms.
Gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction  \citep{16571137}
or automatic detection of heartbeats~\citep{7043102}.
Let us first focus on hidden Markov model (HMM). 
Suppose you have a sequence of observations e.g. amino acids. What is of interest is understanding underlying cause of their occurrence. 
HMM aims to answer that by assuming a specific yet flexible structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
These are "the cause" for the observations. At every step, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

@@

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=20pt,mirror,raise=10pt},yshift=-10pt]
(1,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};

\end{tikzpicture}
\caption{Hidden semi-Markov Model scheme}
\label{fig:hsmm}
\end{figure}

The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
any given hidden state duration distribution. Scheme is shown in figure~\ref{fig:hsmm}.
In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
where $D$ is maximal allowed duration.
As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
higher than with the regular HMM.

The advantage of HSMM model is not only a better performance but also, as 
model structure (see figure~\ref{fig:hsmm}) is straightforward, easier interpretation 
for researcher without mathematical background.

@@

      %\draw [decorate,line width=3pt,decoration={brace,amplitude=10pt}] ([xshift=5mm, yshift=5mm]kmer1) -- ([yshift=-25mm]kmer3) node [black,midway,xshift=0.6cm] {\footnotesize $P_1$};
      \draw[decoration={brace,amplitude=10pt, raise=5pt},decorate, line width=2pt]  
      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{yellow!50!black!90} k-mer};
    \end{tikzpicture} }
    \caption{Scheme of signalHsmm extended wtich the n-gram cleavage site model.}
    \label{fig:ngramext}
    \end{minipage}
\end{figure}

@@

HSMM can be further extended to use some specific motives that might occur in proximity of cleavage site.
Such an extension requires defining additional hidden states. 
To find such motives we analyze n-grams (k-mers) -- vectors of n characters derived from input sequences. 
The rigid structure of n-grams may mirror cleavage sites, which are more 
conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
Scheme of extended HSMM can be seen in figure~\ref{fig:model_hsmm_kMer}.

>>>>

\subsection*{Hidden semi-Markov model}
Hidden semi-Markov model (HSMM) is an extension of hidden Markov model (HMM).
As such, it proved to yield better performance in many fields of bioinformatics including
gene recognition~\citep{Pachter02applicationsof}, protein secondary structure prediction~\citep{16571137}
or automatic detection of heartbeats~\citep{7043102}.
Let us first briefly describe the idea that lies behind HMM. 
Suppose you have a sequence of observations e.g. amino acids, and you are interested in understanding underlying cause of their occurrence. 
HMM aims to answer that question by assuming a specific, yet flexible, structure of the problem.
HMM consists of two stochastic processes. First is discrete Markov chain $X_{t=1}^T$ on the set of so called hidden states $\{S_1, \dots S_n\}$.
They are "the cause" of the observations. At every step $t$, hidden state might change according to transition matrix
$A= (a)_{i,j=1}^n$, where $a_{i,j} = \mathcal{P}(X_{t+1} = S_j | X_t = S_i)$. In our application hidden states are signal peptide regions.
Second process $E_{t=1}^T$ is observation process, defined on the set of possible observations $\{O_1, \dots, O_m\}$. 
Observations are assumed to occur independently. Their distribution depends only on hidden state which emits it. 
Probabilities are given by a matrix $B$, $b_{i,k} = \mathcal{P}(E_t = O_k | X_t = S_i)$.
In our case, observations are (degenerated) amino acids.
Main goal in signalHsmm is to find, for a given peptide, most probable regions boundaries. This is achieved with
Viterbi algorithm.
For a good reference on HMM see~\citep{1989rabinera}.

@@

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
(1,-1.8) -- (5,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_1$}};

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=10pt,mirror,raise=10pt},yshift=-10pt]
(8,-1.8) -- (12,-1.8) node [branch2, midway, yshift=-1cm,color=black] {\textbf{Duration of length $d_k$}};

\draw [color=red!30!black!50, line width=1.5mm, decorate,decoration={brace,amplitude=20pt,mirror,raise=10pt},yshift=-10pt]
(1,-2.8) -- (12,-2.8) node [branch2, midway, yshift=-1.5cm,color=black] {\textbf{Observations, $\sum_{i=1}^k d_i = T$}};

\end{tikzpicture}
\caption{General scheme of hidden semi-Markov model}
\label{fig:hsmm}
\end{figure}

The model we use is Hidden semi-Markov Model(HSMM)~\citep{Yu2010215}. It extends HMM by allowing 
any given hidden state duration distribution. Diagram of HSMM is presented in figure~\ref{fig:hsmm}.
In addition to matrices $A$ and $B$, model is given by probabilities of duration length in hidden states.
$$\mathcal{P}(\text{duration in state} = d | \text{state is } S_i), \;\; i = 1, \dots, n, \;\; d = 1, \dots, D$$
where $D$ is maximum allowed duration.
As our datasets are of reasonable size, $D$ is small -- around 30, computational effort is not much 
higher than with the regular HMM.

The model we use has a very specific structure. Hidden states are signal peptide regions.
Almost all entries in transition matrix $A$ are zeros, because regions are sequential.
Possible transitions are depicted as arrows in figure~\ref{fig:signalHsmm}.
Probabilities of observations for hidden states and hidden states durations were estimated from training data.
The advantage of HSMM model is not only a better performance but also its straightforwardness.
Figure~\ref{fig:signalHsmm} is easy to interpret for a researcher without any mathematical background.

@@

      %\draw [decorate,line width=3pt,decoration={brace,amplitude=10pt}] ([xshift=5mm, yshift=5mm]kmer1) -- ([yshift=-25mm]kmer3) node [black,midway,xshift=0.6cm] {\footnotesize $P_1$};
      \draw[decoration={brace,amplitude=10pt, raise=5pt},decorate, line width=2pt]  
      ([yshift=1mm]kmer1.east) -- ([yshift=-5mm]kmer3.east) node [black,midway,xshift=0.6cm] {\large \color{yellow!50!black!90} k-mer};
    \end{tikzpicture} }
    \caption{The diagram of signalHsmm extended with the n-gram cleavage site model.}
    \label{fig:ngramext}
    \end{minipage}
\end{figure}

@@

HSMM can be further extended to use some specific motives that might occur in proximity of cleavage site.
Such an extension requires defining additional hidden states. 
To find such motives we analyze n-grams (k-mers) -- vectors of n characters derived from input sequences. 
The rigid structure of n-grams may mirror cleavage sites, which are more 
conservative than other parts of signal peptide~\citep{2004hillerpredisi}.
Scheme of extended HSMM can be seen in figure~\ref{fig:ngramext}.
